Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2024-11-19 src/tex/ms.tex
[showyourwork] overleaf sync

Following \citet{Philcox_2019}, to avoid unrealistic star formation histories (that are very `bursty' for early stars), we additionally require that the SFR (parametrized by a $\Gamma$ distribution with shape parameter $a=2$) at the maximum possible stellar birth-time ($13.8$\,Gyr) should be at least 5\% of the mean SFR, ensuring that there is still a reasonable chance of forming a star at this time-step. This corresponds to the constraint $\log_{10}\left(\mathrm{SFR}_\mathrm{peak}\right)>0.294$. For this reason, a truncated Normal prior will be used for the SFR parameter. Furthermore, we constrain $T_i$ to the interval $[1,13.8]$\,Gyr (assuming an age of the Universe of 13.8\,Gyr), ignoring any stars formed before $1$\,Gyr, which is justified as these are expected to be rare.
%
\begin{tiny}
\begin{table*}
\begin{minipage}{\textwidth}
\begin{center}
\caption{Free \texttt{Chempy} parameters for each star, with their prior values and Gaussian widths. Stellar birth-times are set for each star individually from a Uniform prior, based on realistic age estimates.}
\begin{tabularx}{\textwidth}{ >{\raggedleft}p{2.2cm}p{6.5cm}|c{1.9cm}c{2.cm} }
Parameter & Description & $\overline{\theta}_\mathrm{prior}\pm\sigma_\mathrm{prior}$ & Prior from: \\

@@

\label{tab:priors}
\end{tabularx}
\end{center}
\end{minipage}
\end{table*}
\end{tiny}
%
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
%To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{Chempy}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{Chempy} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this are examined in Sec.\,\ref{subsec: mocks_wrong_yield} by performing inference using an alternative set of yields. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully chose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{1998A&A...334..505P} and do not allow for any `hypernovae' (in contrary to \citetalias{2018ApJ...861...40P}).
%
% \begin{table}[]
% \caption{Nucleosynthetic yield tables used in this analysis, matching those of the TNG simulation \citep[Tab.\,2]{2018MNRAS.473.4077P}.}
%     \centering
%     \begin{tabular}{c|c}
%       Type & Yield Table \\
%        \hline
%         SN\,Ia & \citet{1997NuPhA.621..467N}\\
%         SN\,II & \citet{2006ApJ...653.1145K,portinari}\\
%         AGB & \citet{2010MNRAS.403.1413K,2014MNRAS.437..195D};\\
%         & \citet{2014ApJ...797...44F}
%     \end{tabular}
% \label{tab:chempy_TNG_yields}
% \end{table}
%
%TNG only tracks nine elements in their analysis: C, Fe, H, He, Mg, N, Ne, O and Si, reporting the mass-fractions of each \citep{2018MNRAS.473.4077P}. 
In our analysis we only track nine elements: C, Fe, H, He, Mg, N, Ne, O and Si and principally compare the logarithmic abundances [X/Fe] and [Fe/H] (defined by 
%\begin{equation}
    $[\mathrm{X}/\mathrm{Y}] = \log_{10}(N_\mathrm X/N_\mathrm Y)_\mathrm{star} - \log_{10}(N_\mathrm X/N_\mathrm Y)_\odot$
%\end{equation}
for number fraction $N_\mathrm X$ of element X), where $\odot$ denotes the solar number fractions of \citet{2009ARA&A..47..481A}. This uses H for normalization, thus we are left with $n_\mathrm{el}=8$ independent elements which must be tracked by \texttt{Chempy}.%\footnote{In observational contexts, it may be more appropriate to compute abundances relative to Mg rather than Fe (as in \citep{2019ApJ...874..102W}) since Mg is only significantly produced by SN\,II and hence a simpler tracer of chemical enrichment.} 
%In this paper, \texttt{Chempy} will be used as the principal GCE model, which, with the modifications described above, allows for fast prediction of TNG-like chemical abundances for a given set of galactic parameters. It is important to note that the two GCE models have very different parametrizations of galactic physics, with TNG including vastly more effects, thus it is not certain \textit{a priori} how useful \texttt{Chempy} will be in emulating the TNG simulation, although its utility was partially demonstrated in \citet{2018ApJ...861...40P}. This is a necessary test to prepare for an inference on real data.

>>>>

Following \citet{Philcox_2019}, to avoid unrealistic star formation histories (that are very `bursty' for early stars), we additionally require that the SFR (parametrized by a $\Gamma$ distribution with shape parameter $a=2$) at the maximum possible stellar birth-time ($13.8$\,Gyr) should be at least 5\% of the mean SFR, ensuring that there is still a reasonable chance of forming a star at this time-step. This corresponds to the constraint $\log_{10}\left(\mathrm{SFR}_\mathrm{peak}\right)>0.294$. For this reason, a truncated Normal prior will be used for the SFR parameter. Furthermore, we constrain $T_i$ to the interval $[1,13.8]$\,Gyr (assuming an age of the Universe of 13.8\,Gyr), ignoring any stars formed before $1$\,Gyr, which is justified as these are expected to be rare.
%

\begin{tiny}
\begin{table*}
\begin{minipage}{\textwidth}
\begin{center}
\caption{Free \texttt{Chempy} parameters for each star, with their prior values and Gaussian widths. Stellar birth-times are set for each star individually from a Uniform prior, based on realistic age estimates.}
\begin{tabularx}{\textwidth}{ >{\raggedleft}p{2.2cm}p{6.5cm}|c c }
Parameter & Description & $\overline{\theta}_\mathrm{prior}\pm\sigma_\mathrm{prior}$ & Prior from: \\

@@

\label{tab:priors}
\end{tabularx}
\end{center}
\end{minipage}
\end{table*}
\end{tiny}

%
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
%To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{Chempy}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{Chempy} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this are examined in Sec.\,\ref{subsec: mocks_wrong_yield} by performing inference using an alternative set of yields. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully chose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{1998A&A...334..505P} and do not allow for any `hypernovae' (in contrary to \citetalias{2018ApJ...861...40P}).
%
% \begin{table}[]
% \caption{Nucleosynthetic yield tables used in this analysis, matching those of the TNG simulation \citep[Tab.\,2]{2018MNRAS.473.4077P}.}
%     \centering
%     \begin{tabular}{c|c}
%       Type & Yield Table \\
%        \hline
%         SN\,Ia & \citet{1997NuPhA.621..467N}\\
%         SN\,II & \citet{2006ApJ...653.1145K,portinari}\\
%         AGB & \citet{2010MNRAS.403.1413K,2014MNRAS.437..195D};\\
%         & \citet{2014ApJ...797...44F}
%     \end{tabular}
% \label{tab:chempy_TNG_yields}
% \end{table}
%
%TNG only tracks nine elements in their analysis: C, Fe, H, He, Mg, N, Ne, O and Si, reporting the mass-fractions of each \citep{2018MNRAS.473.4077P}. 
In our analysis we only track nine elements: C, Fe, H, He, Mg, N, Ne, O and Si and principally compare the logarithmic abundances [X/Fe] and [Fe/H] (defined by 
%\begin{equation}
    $[\mathrm{X}/\mathrm{Y}] = \log_{10}(N_\mathrm X/N_\mathrm Y)_\mathrm{star} - \log_{10}(N_\mathrm X/N_\mathrm Y)_\odot$
%\end{equation}
for number fraction $N_\mathrm X$ of element X), where $\odot$ denotes the solar number fractions of \citet{2009ARA&A..47..481A}. This uses H for normalization, thus we are left with $n_\mathrm{el}=8$ independent elements which must be tracked by \texttt{Chempy}.%\footnote{In observational contexts, it may be more appropriate to compute abundances relative to Mg rather than Fe (as in \citep{2019ApJ...874..102W}) since Mg is only significantly produced by SN\,II and hence a simpler tracer of chemical enrichment.} 
%In this paper, \texttt{Chempy} will be used as the principal GCE model, which, with the modifications described above, allows for fast prediction of TNG-like chemical abundances for a given set of galactic parameters. It is important to note that the two GCE models have very different parametrizations of galactic physics, with TNG including vastly more effects, thus it is not certain \textit{a priori} how useful \texttt{Chempy} will be in emulating the TNG simulation, although its utility was partially demonstrated in \citet{2018ApJ...861...40P}. This is a necessary test to prepare for an inference on real data.
